---
title: "Take-home Exercise 1: Geospatial Analytics for Public Good"
date: "23 November 2023"
date-modified: "last-modified"
format: html
execute:
  echo: true #all the codes will appear
  eval: true #all the codes will run
  warning: false #dont display if there are any warnings
editor: visual
---

# 1 Overview

## Context

## Objectives

# 2 Getting Started

Ensure that **spdep**, **sf**, **tmap** and **tidyverse** packages are installed:

-   sf is use for importing and handling geospatial data in R,
-   tidyverse is mainly use for wrangling attribute data in R,
-   spdep will be used to compute spatial weights, global and local spatial autocorrelation statistics, and
-   tmap will be used to prepare cartographic quality chropleth map.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)

# -   Creates a package list containing the necessary R packages
# -   Checks if the R packages in the package list have been installed
# -   If not installed, will install the missing packages & launch into R environment.
```

# 3 **Data Preparation**

## The Data

## Import & Preparation

:::panel-tabset
##Aspatial

Firstly, we will import the *Passenger Volume by Origin Destination Bus Stops* dataset downloaded from LTA DataMall by using `read_csv()` or **readr** package.

```{r}
odbus <- read_csv("data/aspatial/origin_destination_bus_202308.csv")
```
## Geospatial

:::
