---
title: "Hands-on Exercise 1: Geospatial Data Wrangling with R"
---

## Overview

In this hands-on exercise, I learn how to import and wrangle geospatial data using appropriate R packages.

## Getting Started

The code chunk below installs and loads sf and tidyverse packages into R environment.

```{r}
pacman::p_load(sf, tidyverse)
```

## Importing Geospatial Data

### Importing polygon feature data in shapefile format

-   *st_read()* function of sf package:

    -   read simple features form file/database, or retrieve layer names and their geometry types(s)

    -   imports `MP14_SUBZONE_WEB_PL` shapefile into R as polygon feature data frame

    -   `dsn` = defines data path; `layer` = provide the shapefile name

```{r}
mpsz = st_read(dsn = "data/geospatial", 
                  layer = "MP14_SUBZONE_WEB_PL")
```

-   Interpretation:

    -   Geospatial objects = multipolygon features

    -   323 multipolygon features and 15 fields in `mpsz` simple feature data frame

    -   `mpsz` in svy21 projected coordinates systems

    -   Bounding box provides x extend and y extend of the data

## Importing polyline feature data in shapefile form

-   Imports CyclingPath shapefile into R as line feature data frame

```{r}
cyclingpath = st_read(dsn = "data/geospatial", 
                         layer = "CyclingPathGazette")
```
